---
title: "project"
output: html_document
date: "2024-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
serie=window(ts(read.table("pasajebcn.dat"),start=1997,freq=12))

plot(serie, main="Passengers in the port of Barcelona.", ylab="Thousand of passengers")
abline(v=1997:2020,lty=3,col=4)
```
```{r}
print(round(serie,0))
```

# 1. Identification: 

# a) Determine the needed transformations to make the series stationary. Justify the transformations
# carried out using graphical and numerical results. 


```{r}
(m<-apply(matrix(serie, nr=12),2,mean))
(v<-apply(matrix(serie, nr=12),2,var))
plot(v~m)
```

```{r}
boxplot(serie~floor(time(serie)))
```
```{r}
lnserie = log(serie)
plot(lnserie)
```

```{r}
(m<-apply(matrix(lnserie, nr=12),2,mean))
(v<-apply(matrix(lnserie, nr=12),2,var))
plot(v~m)
boxplot(lnserie~floor(time(serie)))
```

```{r}
print("Variance of origin time series:")
print(var(serie))
print("Var of ln series:")
print(var(lnserie))
if (var(serie)-var(lnserie) > 0){
  print('Variance decreased. Transformation kept')
}
```

```{r}
monthplot(lnserie)
ts.plot(matrix(lnserie, nr=12))
```

```{r}
d12lnserie = diff(lnserie, 12)
plot(d12lnserie)
abline(h=0)
```

```{r}
monthplot(d12lnserie)
ts.plot(matrix(d12lnserie, nr=12))
```
```{r}
print("Difference of var diff 12 with var of ln series:")
print(var(lnserie)-var(d12lnserie))
if (var(lnserie)-var(d12lnserie) > 0){
  print('Variance decreased. Transformation kept')
}
```


```{r}
d1d12lnserie <- diff(d12lnserie, lag=1)
plot(d1d12lnserie)
abline(h=0)
```


```{r}
print("Difference of var diff 12 with var of diff1 and diff12 series:")
print(var(d12lnserie)-var(d1d12lnserie))
if (var(d12lnserie)-var(d1d12lnserie) > 0){
  print('Variance decreased. Transformation kept')
}
```
# Variance increased so the diff(1) is not kept

# Final: Wt=(1-B^12)log(Xt)


# b) Analyze the ACF and PACF of the stationary series to identify at least two plausible models. Reason
# about what features of the correlograms you use to identify these models. 


```{r}
par(mfrow=c(1,2))
acf(d12lnserie,ylim=c(-1,1),col=c(2,rep(1,11)),lwd=2,lag.max=72)
pacf(d12lnserie,ylim=c(-1,1),col=c(rep(1,11),2),lwd=2,lag.max=72)
par(mfrow=c(1,1))
```
# From PACF -> AR(p) = AR(2), AR(P)=AR(1)
# Model from ACF -> MA(q) = MA(5) or MA(8). We reject that. for MA(Q) = MA(1)
# Since ACF has decreasing lags and then alternated, it is better to select a model from PACF
# d,D -> (0,1)
# 1st model: (S)ARIMA(2,0,0)(1,1,0)12
# 2nd model: (S)ARIMA(2,0,0)(0,1,1)12



# 2. Estimation:
#  a) Use R to estimate the identified models. 

```{r}
(mod1=arima(lnserie,order=c(2,0,0),seasonal=list(order=c(1,1,0),period=12)))

(mod2=arima(lnserie,order=c(2,0,0),seasonal=list(order=c(0,1,1),period=12)))
```

# 3. Validation:
# a) Perform the complete analysis of residuals, justifying all assumptions made. Use the corresponding
# tests and graphical results.
# b) Include analysis of the expressions of the AR and MA infinite models, discuss if they are causal and/or invertible and report some adequacy measures.

```{r}
#################Validation#################################
validation=function(model){
  s=frequency(get(model$series))
  resi=model$residuals
  par(mfrow=c(2,2),mar=c(3,3,3,3))
  #Residuals plot
  plot(resi,main="Residuals")
  abline(h=0)
  abline(h=c(-3*sd(resi),3*sd(resi)),lty=3,col=4)
  #Square Root of absolute values of residuals (Homocedasticity)
  scatter.smooth(sqrt(abs(resi)),main="Square Root of Absolute residuals",
                 lpars=list(col=2))
  
  #Normal plot of residuals
  qqnorm(resi)
  qqline(resi,col=2,lwd=2)
  
  ##Histogram of residuals with normal curve
  hist(resi,breaks=20,freq=FALSE)
  curve(dnorm(x,mean=mean(resi),sd=sd(resi)),col=2,add=T)
  
  
  #ACF & PACF of residuals
  par(mfrow=c(1,2))
  acf(resi,ylim=c(-1,1),lag.max=60,col=c(2,rep(1,s-1)),lwd=2)
  pacf(resi,ylim=c(-1,1),lag.max=60,col=c(rep(1,s-1),2),lwd=2)
  par(mfrow=c(1,1))
  
  #Ljung-Box p-values
  par(mar=c(2,2,1,1))
  tsdiag(model,gof.lag=7*s)
  cat("\n--------------------------------------------------------------------\n")
  print(model)
  
  #Stationary and Invertible
  cat("\nModul of AR Characteristic polynomial Roots: ", 
      Mod(polyroot(c(1,-model$model$phi))),"\n")
  cat("\nModul of MA Characteristic polynomial Roots: ",
      Mod(polyroot(c(1,model$model$theta))),"\n")
  
  suppressMessages(require(forecast,quietly=TRUE,warn.conflicts=FALSE))
  plot(model)
  
  #Model expressed as an MA infinity (psi-weights)
  psis=ARMAtoMA(ar=model$model$phi,ma=model$model$theta,lag.max=36)
  names(psis)=paste("psi",1:36)
  cat("\nPsi-weights (MA(inf))\n")
  cat("\n--------------------\n")
  print(psis[1:24])
  
  #Model expressed as an AR infinity (pi-weights)
  pis=-ARMAtoMA(ar=-model$model$theta,ma=-model$model$phi,lag.max=36)
  names(pis)=paste("pi",1:36)
  cat("\nPi-weights (AR(inf))\n")
  cat("\n--------------------\n")
  print(pis[1:24])
   
  cat("\nDescriptive Statistics for the Residuals\n")
  cat("\n----------------------------------------\n") 
  
  suppressMessages(require(fBasics,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  # print(basicStats(resi))
  
  ## Add here complementary tests (use with caution!)
  ##---------------------------------------------------------
  cat("\nNormality Tests\n")
  cat("\n--------------------\n")
 
  ##Shapiro-Wilks Normality test
  print(shapiro.test(resi))

  suppressMessages(require(nortest,quietly=TRUE,warn.conflicts=FALSE))
  ##Anderson-Darling test
  print(ad.test(resi))
  
  suppressMessages(require(tseries,quietly=TRUE,warn.conflicts=FALSE))
  ##Jarque-Bera test
  print(jarque.bera.test(resi))
  
  cat("\nHomoscedasticity Test\n")
  cat("\n--------------------\n")
  suppressMessages(require(lmtest,quietly=TRUE,warn.conflicts=FALSE))
  ##Breusch-Pagan test
  obs=get(model$series)
  print(bptest(resi~I(obs-resi)))
  
  cat("\nIndependence Tests\n")
  cat("\n--------------------\n")
  
  ##Durbin-Watson test
  print(dwtest(resi~I(1:length(resi))))
  
  ##Ljung-Box test
  cat("\nLjung-Box test\n")
  print(t(apply(matrix(c(1:4,(1:4)*s)),1,function(el) {
    te=Box.test(resi,type="Ljung-Box",lag=el)
    c(lag=(te$parameter),statistic=te$statistic[[1]],p.value=te$p.value)})))
  
}
```


```{r}
validation(mod1)
```
```{r}
validation(mod2)
```
# c) Check the stability of the proposed models and evaluate their capability of prediction, reserving the last 12 observations.

# create validation series (with and w/o the last 12 months)
```{r}
ultim=c(2018,12)

serie2=window(serie,end=ultim)
lnserie2=log(serie2)
#serie1=window(serie,end=ultim+c(1,0))
#lnserie1=log(serie1)
```

# create the first proposed model (S)ARIMA(2,0,0)(1,1,0)12
```{r}
pdq1=c(2,0,0)
PDQ1=c(1,1,0)

# (modA=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
(mod1=arima(lnserie2,order=pdq1,seasonal=list(order=PDQ1,period=12)))
```
# make predictions with the model
```{r}
pred1=predict(mod1,n.ahead=12)
pr1<-ts(c(tail(lnserie2,1),pred1$pred),start=ultim,freq=12)
se1<-ts(c(0,pred1$se),start=ultim,freq=12)

#Intervals
tl1<-ts(exp(pr1-1.96*se1),start=ultim,freq=12)
tu1<-ts(exp(pr1+1.96*se1),start=ultim,freq=12)
pr1<-ts(exp(pr1),start=ultim,freq=12)


ts.plot(serie,tl1,tu1,pr1,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq1,collapse=","),")(",paste(PDQ1,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```

# calculate model accuracy statistics
```{r}
obs=window(serie,start=ultim+c(0,1))
pr1=window(pr1,start=ultim+c(0,1))
ts(data.frame(LowLim=tl1[-1],Predic=pr1,UpperLim=tu1[-1],Observ=obs,Error=obs-pr1,PercentError=(obs-pr1)/obs),start=ultim+c(0,1),freq=12)
mod1.RMSE=sqrt(sum((obs-pr1)^2)/12)
mod1.MAE=sum(abs(obs-pr1))/12
mod1.RMSPE=sqrt(sum(((obs-pr1)/obs)^2)/12)
mod1.MAPE=sum(abs(obs-pr1)/obs)/12

data.frame("RMSE"=mod1.RMSE,"MAE"=mod1.MAE,"RMSPE"=mod1.RMSPE,"MAPE"=mod1.MAPE)

mCI1=mean(tu1-tl1)

cat("\nMean Length CI: ",mCI1)
```

# do the same for the second model
# 2nd model: (S)ARIMA(2,0,0)(0,1,1)12
```{r}
pdq2=c(2,0,0)
PDQ2=c(0,1,1)

(mod2=arima(lnserie2,order=pdq2,seasonal=list(order=PDQ2,period=12)))
```
# make predictions
```{r}
pred2=predict(mod2,n.ahead=12)
pr2<-ts(c(tail(lnserie2,1),pred2$pred),start=ultim,freq=12)
se2<-ts(c(0,pred2$se),start=ultim,freq=12)

#Intervals
tl2<-ts(exp(pr2-1.96*se2),start=ultim,freq=12)
tu2<-ts(exp(pr2+1.96*se2),start=ultim,freq=12)
pr2<-ts(exp(pr2),start=ultim,freq=12)


ts.plot(serie,tl2,tu2,pr2,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=ultim[1]+c(-2,+2),type="o",main=paste("Model ARIMA(",paste(pdq2,collapse=","),")(",paste(PDQ2,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+2),lty=3,col=4)
```

# calculate statistics
```{r}
obs=window(serie,start=ultim+c(0,1))
pr2=window(pr2,start=ultim+c(0,1))
ts(data.frame(LowLim=tl2[-1],Predic=pr2,UpperLim=tu2[-1],Observ=obs,Error=obs-pr2,PercentError=(obs-pr2)/obs),start=ultim+c(0,1),freq=12)
mod2.RMSE=sqrt(sum((obs-pr2)^2)/12)
mod2.MAE=sum(abs(obs-pr2))/12
mod2.RMSPE=sqrt(sum(((obs-pr2)/obs)^2)/12)
mod2.MAPE=sum(abs(obs-pr2)/obs)/12

data.frame("RMSE"=mod2.RMSE,"MAE"=mod2.MAE,"RMSPE"=mod2.RMSPE,"MAPE"=mod2.MAPE)

mCI2=mean(tu2-tl2)

cat("\nMean Length CI: ",mCI2)
```

# compare the two models based on the calculated metrics + AIC & BIC
```{r}
result=data.frame(
  par=c(length(coef(mod1)),length(coef(mod2))),
  Sigma2Z=c(mod1$sigma2,mod2$sigma2),
  AIC=c(AIC(mod1),AIC(mod2)),
  BIC=c(BIC(mod1),BIC(mod2)),
   RMSE=c(mod1.RMSE,mod2.RMSE),
  MAE=c(mod1.MAE,mod2.MAE),
  RMSPE=c(mod1.RMSPE,mod2.RMSPE),
  MAPE=c(mod1.MAPE,mod2.MAPE),
  meanLength=c(mCI1,mCI2)
  )

row.names(result)=c("ARIMA(2,0,0)(1,1,0)12","ARIMA(2,0,0)(0,1,1)12")

result
```
# Conclusion
# Based on the metrics, model2 outperforms model1

# d) Select the best model for forecasting.

# so we define the final model (trained on all data)
```{r}
ultim=c(2018,12)

serie1=window(serie,end=ultim+c(1,0))
lnserie1=log(serie1)

pdq=c(2,0,0)
PDQ=c(0,1,1)

(modFinal=arima(lnserie1,order=pdq,seasonal=list(order=PDQ,period=12)))
```

# 4. Predictions
# a) Obtain long term forecasts for the twelve months following the last observation available; provide also confidence intervals.

# make the 12 month forecast
```{r}
pred=predict(modFinal,n.ahead=12)
pr<-ts(c(tail(lnserie1,1),pred$pred),start=ultim+c(1,0),freq=12)
se<-ts(c(0,pred$se),start=ultim+c(1,0),freq=12)

tl<-ts(exp(pr-1.96*se),start=ultim+c(1,0),freq=12)
tu<-ts(exp(pr+1.96*se),start=ultim+c(1,0),freq=12)
pr<-ts(exp(pr),start=ultim+c(1,0),freq=12)

ts.plot(serie,tl,tu,pr,lty=c(1,2,2,1),col=c(1,4,4,2),xlim=c(ultim[1]-2,ultim[1]+3),type="o",main=paste("Model ARIMA(",paste(pdq,collapse=","),")(",paste(PDQ,collapse=","),")12",sep=""))
abline(v=(ultim[1]-2):(ultim[1]+3),lty=3,col=4)
```
# print the predictions as well
```{r}
(forecast=window(cbind(tl,pr,tu),start=ultim+c(1,0)))
```

```{r}

```